A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-02 02-ceu.tex
Início da Seção 2 revisado: - tirei "output" - mudei o exemplo para usar "input" e "finalize" - falei sobre o mecanismo ortogonal de abortamento - faltou falar sobre timers, por alguma razão não está em lugar nenhum do artigo (falta de espaço?)

\CEU is a synchronous reactive language inspired by
Esterel~\cite{esterel.ieee91} in which programs advance in a sequence of
discrete reactions to external events.
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
C~calls) execute in no time in accordance to the synchronous
hypothesis~\cite{rp.hypothesis}.
%
The \code{await} is the only \CEU statement that actually ``consumes'' time.
%%
An~\code{await} statement blocks the current trail allowing the program to
advance its other trails; when all trails are blocked, the reaction
terminates and control returns to the environment.

In \CEU, every execution path within a loop must contain at least one
\code{await} statement~\cite{ceu.sensys13,esterel.primer}.
%
This restriction, which is checked statically by the \CEU compiler, ensures
that every reaction runs in bounded time, eventually terminating with all
program trails blocked in \code{await} statements.
%
\CEU has a further restriction which it shares with Esterel and synchronous
languages in general~\cite{esterel.preemption}: computations that actually
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

Listing~\ref{lst.syntax} below shows a compact reference of~\CEU.

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations:}|
input  $\T<type>$ $\T<ids>$;|\C{.6}{declare external input events}|
output $\T<type>$ $\T<ids>$;|\C{.6}{declare external output events}|
event  $\T<type>$ $\T<ids>$;|\C{.6}{declare internal events}|
var    $\T<type>$ $\T<id>$ = $\T<exp>$;|\C{.6}{declare and initialize variable}|

|\C{1.}{Event handling:}|
$\T<id>$ = await $\T<id>$;|\C{.6}{await event and assign the received value}|
$\T<id>$ = await $\T<time>$;|\C{.6}{await time and assign the delayed delta}|
emit $\T<id>$($\T<exp>$);|\C{.6}{emit event passing a value}|

|\C{1.}{Control-flow:}|
$\T<stmt>$ ; $\T<stmt>$|\C{.45}{sequence}|
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end|\C{.45}{conditional}|
loop do $\T<stmts>$ end|\C{.45}{repetition}|
finalize $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{finalization}|

par/or  do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{aborts when any side terminates}|
par/and do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{terminates when all sides terminate}|
par     do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{never terminates}|

|\C{1.}{Assignment \& integration with C:}|
$\T<id>$ = $\T<exp>$;|\C{.45}{assign value to variable}|
_$\T<id>$($\T<exps>$)|\C{.45}{call C function (id starts with `\_'\,)}|
\end{lstlisting}
\egroup

To make matters concrete, consider the program of Listing~\ref{lst.ceu}.
%
This program continuously turns a LED~on for~2 seconds and off for~1 second,
and terminates after~1 minute of activity with the LED off.
%
The implementation uses a \code{par/or} to run two activities in parallel:
an endless loop that blinks the LED on and off (lines~2--7), and a single
statement that waits for~1 minute before terminating (line~9).
%
The \code{par/or} block stands for a \emph{parallel-or} composition; when
executed it creates~$n$ parallel trails (in this case, $n=2$) and rejoins
them when any of these~$n$ trails terminates, automatically aborting the
other trails.

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={\CEU program that blinks a LED during 1 minute.},
  label={lst.ceu},
]
par/or do
    loop do
        _led(1);
        await 2s;
        _led(0);
        await 1s;
    end
with
    await 1min;
end
_led(0);
\end{lstlisting}

In \CEU, any identifier prefixed with an underscore (e.g., \code{_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate synchronously, the compiler environment
should only provide access to~C operations that are assumed run in zero
time, such as non-blocking~I/O and access to \code{struct}'s.

@@

In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
(lines~3--6) that continuously awaits its identifying event (line~4), and
increments the value passed to it by reference (lines~5).
%
A trail in parallel (lines~8--11) invokes the subroutine through the
\code{emit} statement at line~10.
%
Given the stack-based execution mode of internal events, after the emit
statement at line~10 is executed, the calling trail pauses, the subroutine
awakes (line~4), runs its body (yielding \code{v=2}), loops, and awaits the
next ``call'' (line~4, again).
%
Only after this sequence does the calling trail resumes and moves on to
execute the assertion on line~11.

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (line~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same bigger reaction to some external
event.

@@

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}
%
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are concurrent but deterministic.}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

\subsection{Abortion and Finalization}

>>>>

\CEU is a synchronous reactive language in which programs advance in a sequence
of discrete reactions to external events.
%
It is designed for control-intensive applications, supporting concurrent lines
of execution, known as \emph{trails}, and instantaneous broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
system calls) are also instantaneous in accordance to the synchronous
hypothesis~\cite{rp.hypothesis}.
%
\CEU provides an \code{await} statement which is the only that actually
``consumes'' time.
%
An \code{await} statement blocks the current running trail allowing the program
to advance its other trails; when all trails are blocked, the reaction
terminates and control returns to the environment.

In \CEU, every execution path within loops must contain at least one
\code{await} statement~\cite{ceu.sensys13,esterel.primer}.
%
This restriction, which is statically checked by the compiler, ensures that
every reaction runs in bounded time, eventually terminating with all trails
blocked in \code{await} statements.
%
\CEU has an additional restriction, which it shares with Esterel and
synchronous languages in general~\cite{esterel.preemption}: computations that
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

Listing~\ref{lst.syntax} shows a compact reference of~\CEU:

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations}|
input $\T<type>$ $\T<id>$; | \C{.6}{declares external input event} |
event $\T<type>$ $\T<id>$; | \C{.6}{declares internal event}       |
var   $\T<type>$ $\T<id>$; | \C{.6}{declares variable}             |

|\C{1.}{Event handling}|
$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits event and assigns received value} |
$\T<id>$ = await $\T<time>$; | \C{.6}{awaits time and assigns delayed delta}   |
emit $\T<id>$($\T<exp>$);    | \C{.6}{emits event passing value}               |

|\C{1.}{Control flow}|
$\T<stmt>$ ; $\T<stmt>$                       | \C{.45}{sequence}     |
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \C{.45}{conditional}  |
loop do $\T<stmts>$ end                   | \C{.45}{repetition}   |
finalize [$\T<stmts>$] with $\T<stmts>$ end    | \C{.45}{finalization} |

|\C{1.}{Logical parallelism}|
par/or  do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{aborts when any side ends}      |
par/and do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{terminates when all sides ends} |
par     do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{never terminates}               |

|\C{1.}{Assignment \& Integration with C}|
$\T<id>$ = $\T<exp>$; | \C{.6}{assigns value to variable}                |
_$\T<id>$($\T<exps>$) | \C{.6}{calls C function (id starts with `\_'\,)} |
\end{lstlisting}
\egroup

Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED with a
frequency of 1 second, terminating with a button press always with the LED off.
%
The implementation first declares the \code{BUTTON} as an input event (ln.~1).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single statement that waits for a button press before terminating (ln.~3),
and an endless loop that blinks the LED on and off (ln.~8--13).
The \code{finalize} clause (ln.~5--7) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~6).

The \code{par/or} composition, which stands for a \emph{parallel-or}, provides
an orthogonal abortion mechanism in which its composed trails do not know when
and how they are aborted (i.e., abortion is external to them).
%
This is possible to do safely in synchronous languages due to the accurate
control of concurrent activities, i.e., in between every reaction, the whole
system is idle and consistent~\cite{esterel.preemption}.
%
The finalization mechanism extends orthogonal abortion to also work with
activities that use stateful resources from the environment (such as files and
network handlers), as we discuss in Section~\ref{sec.ceu.fin}.
%
\fs{First-class timers.}

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A program in \CEU that blinks a LED every second, terminating on a
           button press in a consistent state.},
  label={lst.ceu},
]
input void BUTTON;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await 1s;
        _led(0);
        await 1s;
    end
end
\end{lstlisting}

In \CEU, any identifier prefixed with an underscore (e.g., \code{_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple accesses to data
structures.

@@

In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
(ln.~3--6) that continuously awaits its identifying event (ln.~4), and
increments the value passed to it by reference (ln.~5).
%
A trail in parallel (ln.~8--11) invokes the subroutine through an \code{emit}
statement (ln.~10).
%
Given the stack-based execution mode of internal events, as the emit executes,
the calling trail pauses, the subroutine awakes (ln.~4), runs its body
(yielding \code{v=2}), loops, and awaits the next ``call'' (ln.~4, again).
%
Only after this sequence does the calling trail resumes and moves on to
execute the assertion (ln.~11).

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (ln.~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same bigger reaction to some external
event.

@@

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}
%
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are concurrent but deterministic.}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

\subsection{Abortion and Finalization}
\label{sec.ceu.fin}
